Red Wine Quality by Joe Lundberg
========================================================
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
knitr::opts_chunk$set(error = TRUE)
```

```{r echo=FALSE, Load_the_Data}
wineInfo <- read.csv('wineQualityReds.csv')
```
> I chose to explore the red wine data set. I am not a wine drinker but have 
always been curious on what makes a good wine. I plan to explore what chemical 
properties make a red wine have a good quality. This data set contains 1599 
red wines with 11 columns on the chemical properties of the wine.

# Univariate Plots Section

> There looks to be roughly 1600 records with 13 columns in the table. The 
record count and column names can be seen below.

```{r echo=FALSE, Univariate_Plots}
NROW(wineInfo)
names(wineInfo)
```
```{r echo=FALSE}
qplot(x= quality, data = wineInfo)
```
> This chart shows that most wines that were looked at were mid range for wine 
quality.

```{r echo=FALSE}
qplot(x= alcohol, data = wineInfo)
```
> This chart shows the count of wine by alcohol percentage.

```{r echo=FALSE}
qplot(x= density, data = wineInfo)
```
> This chart shows the density of wine for the data set.

```{r echo=FALSE}
qplot(x= citric.acid, data = wineInfo)
```

> This chart shows the amount of citric acid in wine for the data set.

```{r echo=FALSE}
qplot(x= fixed.acidity, data = wineInfo)
```


> This chart shows the fixed acidity level in wine for the data set.

```{r echo=FALSE}
qplot(x= volatile.acidity, data = wineInfo)
```

> This chart shows the volatile acidity level in wine for the data set.

```{r echo=FALSE}
qplot(x= chlorides, data = wineInfo)
```

>This chart shows the chloride level in wine for the data set.

### What is the structure of your dataset?

> My dataset consists 1599 samples of wine. There are 11 variables that are 
defined with the wine. All variables are numeric data types with the sample 
and quality being integer data types.

### What is/are the main feature(s) of interest in your dataset?

>I am mainly interested in what variables make up the best quality of wine. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

>I think the amount of alcohol, acids and sugars will help determine what 
makes a good wine.

### Did you create any new variables from existing variables in the dataset?

> I have not created any new variables yet in my analysis.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

> I don't think this will be needed until I start looking at bivariate or
multivarate anaysis.

# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots}
ggplot(aes(x = quality, y= alcohol),data = wineInfo) + geom_point()
```
>This scatter plot compares the quality of a wine by it's alcohol percentage.

```{r echo=FALSE}
ggplot(aes(x = quality, y= citric.acid),data = wineInfo) + geom_point()
```
>This scatter plot compares the quality of a wine by it's amount of citric acid.

```{r echo=FALSE}
ggplot(aes(x = quality, y= residual.sugar),data = wineInfo) + geom_point()
```
>This scatter plot compares the quality of a wine by it's amount of residual
sugar.

```{r echo=FALSE}
ggplot(aes(x = quality, y= density),data = wineInfo) + geom_point()
```

>This scatter plot compares the quality of a wine by it's density.

```{r echo=FALSE}
ggplot(aes(x = residual.sugar, y= alcohol),data = wineInfo) + geom_point()
```

>This scatter plot compares the sugar level of a wine with its alcohol content.

```{r echo=FALSE}
ggplot(aes(x = density, y= pH),data = wineInfo) + geom_point()
```

>This scatter plot compares the density of a wine with its pH level.

```{r echo=FALSE}
ggplot(aes(x = pH, y= citric.acid),data = wineInfo) + geom_point()
```
```{r echo=FALSE}
ggplot(aes(x = pH, y= fixed.acidity),data = wineInfo) + geom_point()
```
```{r echo=FALSE}
ggplot(aes(x = pH, y= volatile.acidity),data = wineInfo) + geom_point()
```
>The last three scatter plots compares the acids of a wine with its pH level.

```{r echo=FALSE}
ggplot(aes(x = quality, y= chlorides),data = wineInfo) + geom_point()
```

>This charts shows the relationship bewtween chlorides and quality.

# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset? 
>It appears that the higher the sugar level usually means lower the alcohol
content. I also noticed the lower levels of acid usually means a higher pH 
level. It also shows the lower the chloride level the higher the quality.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
> I thought the relationship between sugar and quality was very interesting. I
wasn't expecting a lower sugar level meant a better quality of wine but the 
data proves that is true.

### What was the strongest relationship you found?
> I feel that the sugar to alcohol content was a strong relationship. It 
appeared that most wines stuck to a lower amount of sugar and that the alcohol
content was very similar. There were some outliers that led me to believe the 
more sugar added the less the alcohol content.

# Multivariate Plots Section
>In this section I decided to create a new variable call quality.type. By doing
this I am able to categorize the quality of wines into three categories; 
excellent, average and poor. This way I will be able to compare multiple 
variables with the quality type and by doing so I should be able to see results
that will define an excellent wine. Below is the code I used to create the new
variable.



```{r echo=FALSE,results="hide", Multivariate_Plots}
wineInfo$quality.type
attach(wineInfo)
wineInfo$quality.type[quality > 6] <- "Excellent"
wineInfo$quality.type[quality  > 3 & quality <= 6] <- "Average"
wineInfo$quality.type[quality  <= 3] <- "Poor"
#detach(wineInfo)

ggplot(wineInfo, aes(x = alcohol, y = residual.sugar, colour = quality.type)) +
  geom_point() + xlab(quality.type)
```
>This charts shows the relationship bewtween residual sugar and alcohol based
on the quality type. It appears that most average wines are on low end for 
sugar and alcohol. It also looks like a lot of the excellent wines tend to have
higher alcohol and sugar content.

```{r echo=FALSE}

ggplot(wineInfo, aes(x = citric.acid, y = fixed.acidity, colour = quality.type)) +
  geom_point() + xlab(quality.type)
```
>This charts shows the relationship bewtween citric acid and fixed acidity based
on the quality type. It appears that most average wines are on low end for 
both. It looks like the most wines that are excellent have about .5 for citric
acid and 8 to 9 for fixed acidity. 

```{r echo=FALSE}

ggplot(wineInfo, aes(x = citric.acid, y = volatile.acidity, colour = quality.type)) +
  geom_point() + xlab(quality.type)
```

>This charts shows the relationship bewtween citric acid and volatile acidity 
based on the quality type. It appears that most average wines are on low end 
for citric acid and on the mid to high range for volatile acidity. It looks 
like the most wines that are excellent have about .5 for citric
acid and are low for the volatile acidity level. 

```{r echo=FALSE}

ggplot(wineInfo, aes(x = pH, y = alcohol, colour = quality.type)) +
  geom_point() + xlab(quality.type)
```
>This charts shows the relationship bewtween alcohol and the pH level based on
the quality type. It appears all the wines tested are on the acidic side of the
pH scale. This tells me that is a common characteristic of wine. I also see the 
most excellent are higher in alcohol content and I don't feel pH has much 
correlation in determining a wines quality.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

>I believe from what I found that part of making an excellent wine is keeping
the sugar levels low, the alcohol percentage in the middle range and having a
low volatile acidity level as well as around .5 for citric acid.

### Were there any interesting or surprising interactions between features?

>I was suprised that the higher a fixed acidity level the lower quality of wine.
I then researched this further and found out that this can make wine taste like
vinegar.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
QualityPlot <- ggplot(wineInfo,aes(quality))+
  geom_histogram(binwidth=1,boundary=-0.5)
print(QualityPlot + ggtitle("Quality Count"))
```

### Description One
>I chose this chart because it really captures the picture of the size of the
data set being used. As you can see most wines that were analysed were already 
mid range wines. It would be nice to have a larger data set that included more
excellent and poor wines. This data was also what I used to make my quality.type
variable. 

### Plot Two
```{r echo=FALSE, Plot_Two}
Vacidity <- ggplot(wineInfo, aes(x = citric.acid, y = volatile.acidity, colour = quality.type)) +
  geom_point() + xlab(quality.type)
print(Vacidity + ggtitle("Volatile Acidity by Quality Type"))

```

### Description Two

> I chose this chart because it made me dig into what the volatile acidity of 
wine can do and found out that it can leave a bad taste to the wine. I think 
this is a very important variable to look at when judging a wine. 

### Plot Three
```{r echo=FALSE, Plot_Three}
SugarLevel <-ggplot(wineInfo, aes(x = alcohol, y = residual.sugar, colour = quality.type)) +
  geom_point() + xlab(quality.type)
print(SugarLevel + ggtitle("Sugar Level by Quality Type"))
```

### Description Three

> I chose this chart because I was really interested in the correlation between
sugar level and alcohol percentage. I think it really represents that an
excellent wine has higher alcohol percentage and is on the lower end for sugar
content.

------

# Reflection

> The red wines data set contains information on 1599 wines with 12 variables.
I started by digging into each variables in the data set. From there it lead me
into looking further into sugar levels, alcohol levels, citric acid levels and
acidity both volatile and fixed. I am not a wine drinker so my biggest question 
was what makes an excellent wine and if I found this answer maybe that would be 
the wine for me.
I do feel the data set is small and most wines on the list were average quality 
wines. This could skew my assumptions as to what an excellent wine is. I feel 
I found what an excellent constists of. It appears that it should be lower in 
sugar, mid alcohol content and mid citric acid. The volatile and fixed acidity 
should also be low to eliminate the vinegar taste. 